How did the introduction of Stemmed Points affect Mobility and Site Occupation during the Late Pleistocene in Korea?

### 2021-04-12 analysis/paper/paper.Rmd
add some inline R code

After the first excavation of a Paleolithic site in Korea at Seokjang-ri site in the 1960s, more than 200 Paleolithic sites have been discovered in South Korea [@lee2019were]. We selected sites dated to before and after the transition period (42-35 ka) to analyze the process of technological change. This resulted in a sample of 33 assemblages from 22 sites spanning 49-24 ka (Figure \@ref(fig:korea-map)). We identified multiple assemblages in a site where culturally sterile deposits separated artifact-bearing deposits, or where stratigraphic units could be identified by major differences in the texture and composition of the sedimentary deposit containing the stone artifacts. For example, the Hwadaeri site has three cultural layers. The lowest horizon at the bottom, dated to 39,000±1400 BP by OSL, contains coarse flake tools made of vein quartz and quartzite. The middle cultural horizon has stemmed points made of porphyry, with quartz and quartzite dominating the assemblage. This layer was dated to 31,200±900 BP by radiocarbon dating and 30,000±1,700 BP by OSL. The uppermost layer, dated to 22,000±100 BP by OSL, contains blades, scrapers, awls, and denticulates [@seong2009emergence]. We obtained the data from published excavation reports held by provincial museums.

@@

@@

We used radiocarbon dates as a proxy for human population to infer the demographic context of technological change. Summed probability distributions (SPD) of radiocarbon dates are increasingly used to infer temporal trends in past human populations [@rick1987dates; @bamforth2012radiocarbon; @shennan2013regional; @timpson2014reconstructing; @contreras2014summed; @riris2019widespread]. Despite concerns about limitations of the validity of SPD [@williams2012use; @bamforth2012radiocarbon; @carleton2021evaluating], several studies have shown good agreement between SPDs and other archaeological indicators (e.g. site counts, settlement size) so this method may allow a first approximation of population fluctuations [@french2015upper; @palmisano2017comparing]. Although this is often used with samples of thousands of ages, this method can also be useful for small sample sizes. For instance, @timpson2014reconstructing apply SPD to 93 ages from Eastern Middle Sweden and demonstrate that the result is equivalent to a larger sample size (n= 243). We used 108 radiocarbon ages from 22 sites to examine connections between population trends during the Late Pleistocene in Korea and technological innovation.

@@

Figure \@ref(fig:site-elevation) shows that people occupied sites located at similar elevations, around 0 to 200 meters above sea level. As we saw for artifact density, there is an increase in the upper bound of site elevation in more recent times. The most striking pattern in the elevation data is the substantial difference in elevation between sites with stemmed points and sites without. The distribution of elevations of sites without stemmed points is generally lower than the elevations of sites with stemmed points. Although not a statistically significant difference in elevations (p = 0.102), this may indicate that forager groups who used stemmed points were generally more able to occupy higher altitudes, while the groups without stemmed points tended to prefer lower altitudes. According to our locally weighted regression, sites without stemmed points reach a maximum elevation of about 150 m at around 39 ka. Sites containing stemmed points reach a maximum elevation of 175 m at 28 ka.

@@

Among the sites in our sample there is a variation of about 5 degrees in the mean annual temperature (MAT), mostly controlled by elevation (Figure \@ref(fig:climate-model)). Through MIS 3, the temperature gradually decreased until the Last Glacial Maximum (LGM, 26.5-20 ka) (Clark et al., 2009). The MAT of MIS 2, including LGM, was relatively stable. The temperature increased again from late MIS 2 towards MIS 1. Compared to the east side of the Korean Peninsula, the west side tends to be relatively warmer. The range of MAT at the sites over our study period is 2-10℃. Gihwari and Gigok sites have the lowest (3℃) and Jeungsan-ri site has the highest MAT (8℃). The mean MAT for all sites fluctuated within 4℃ between 50 ka and 10 ka. The first appearance of stemmed points occurred in the middle of the decreasing MAT trend in MIS 3, at 40-35 ka. Figure 7D shows a negative relationship between temperature and elevation (p = 0.073). For example, Gihwari Cave site, with one of the lowest MAT distributions, is located in the highest elevation around 300m above sea level. Jeungsanri is located at a lower elevation and has one of the highest MAT distributions.

@@

@@

Figure \@ref(fig:spd-model) shows the summed probability distribution (SPD) of the radiocarbon ages from the sites in our sample overlaid with distributions (indicated by the grey envelopes) generated from four models of demographic change for the period 50 ka to 10 ka (exponential, uniform, linear, and logistic). The statistical significance of each of the models we fit to the observed SPD was computed using 200 simulations, and the global p-values are 0.004, 0.004, 0.009, and 0.004, respectively. These low p-values indicates significant deviations in the observed SPD from the modeled distributions. Overall, the linear model is the best fit with the observed SPD, having the lowest AIC value of the three models where we could compute this. We can say that the population did not grow or decline strongly logistically, exponentially or linearly during the Late Pleistocene, but instead followed a noisy distribution.  The most striking results here are the SPD showing a steady growth rising to a peak at about 35 ka, indicated by the significant deviations from all four null models (shaded in red). The SPD remains high with minor deviations until declining to a lower level at 32 ka. After 32 ka the SPD remains generally low, with frequent temporary fluctuations.

```{r spd-model, fig.cap="Summed probability distribution of 108 radiocarbon dates. The black solid line represents actual radiocarbon ages. The grey shaded region shows the Monte Carlo envelope that encompasses the 95% confidence interval for the null models. The red and blue vertical bands highlight the portions of the SPD where positive and negative deviations are detected. We applied logistic, exponential and linear to AIC along with the logistical growth model (fit) to evaluate each model [@riris2019widespread]. We excluded the uniform model because it has no fitted curve or surface to evaluate. The best fit model has the lowest AIC score. In this case, the linear model is the best fit for the observed data", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/008-summed-probability-distributions.R"))}
quietly(source(here::here("analysis/paper/008-summed-probability-distributions.R")))
knitr::include_graphics(here::here("analysis/figures/008-summed-probability-distribution-models.png"))
```

>>>>

@@

After the first excavation of a Paleolithic site in Korea at Seokjang-ri site in the 1960s, more than 200 Paleolithic sites have been discovered in South Korea [@lee2019were]. We selected sites dated to before and after the transition period (42-35 ka) to analyze the process of technological change. This resulted in a sample of 33 assemblages from `r nrow(korean_archaeological_site_locations)` sites spanning 49-24 ka (Figure \@ref(fig:korea-map)). We identified multiple assemblages in a site where culturally sterile deposits separated artifact-bearing deposits, or where stratigraphic units could be identified by major differences in the texture and composition of the sedimentary deposit containing the stone artifacts. For example, the Hwadaeri site has three cultural layers. The lowest horizon at the bottom, dated to 39,000±1400 BP by OSL, contains coarse flake tools made of vein quartz and quartzite. The middle cultural horizon has stemmed points made of porphyry, with quartz and quartzite dominating the assemblage. This layer was dated to 31,200±900 BP by radiocarbon dating and 30,000±1,700 BP by OSL. The uppermost layer, dated to 22,000±100 BP by OSL, contains blades, scrapers, awls, and denticulates [@seong2009emergence]. We obtained the data from published excavation reports held by provincial museums.

@@

```{r spd-model-prep,, cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/008-summed-probability-distributions.R"))}
quietly(source(here::here("analysis/paper/008-summed-probability-distributions.R")))
```

We used radiocarbon dates as a proxy for human population to infer the demographic context of technological change. Summed probability distributions (SPD) of radiocarbon dates are increasingly used to infer temporal trends in past human populations [@rick1987dates; @bamforth2012radiocarbon; @shennan2013regional; @timpson2014reconstructing; @contreras2014summed; @riris2019widespread]. Despite concerns about limitations of the validity of SPD [@williams2012use; @bamforth2012radiocarbon; @carleton2021evaluating], several studies have shown good agreement between SPDs and other archaeological indicators (e.g. site counts, settlement size) so this method may allow a first approximation of population fluctuations [@french2015upper; @palmisano2017comparing]. Although this is often used with samples of thousands of ages, this method can also be useful for small sample sizes. For instance, @timpson2014reconstructing apply SPD to 93 ages from Eastern Middle Sweden and demonstrate that the result is equivalent to a larger sample size (n = 243). We used `r nrow(dates_clean)` radiocarbon ages from `r nrow(korean_archaeological_site_locations)` sites to examine connections between population trends during the Late Pleistocene in Korea and technological innovation.

@@

@@

Figure \@ref(fig:site-elevation) shows that people occupied sites located at similar elevations, around 0 to 200 meters above sea level. As we saw for artifact density, there is an increase in the upper bound of site elevation in more recent times. The most striking pattern in the elevation data is the substantial difference in elevation between sites with stemmed points and sites without. The distribution of elevations of sites without stemmed points is generally lower than the elevations of sites with stemmed points. Although not a statistically significant difference in elevations (`r apa::apa(elev_sp_ttest)`), this may indicate that forager groups who used stemmed points were generally more able to occupy higher altitudes, while the groups without stemmed points tended to prefer lower altitudes. According to our locally weighted regression, sites without stemmed points reach a maximum elevation of about 150 m at around 39 ka. Sites containing stemmed points reach a maximum elevation of 175 m at 28 ka.

@@

Among the sites in our sample there is a variation of about 5 degrees in the mean annual temperature (MAT), mostly controlled by elevation (Figure \@ref(fig:climate-model)). Through MIS 3, the temperature gradually decreased until the Last Glacial Maximum (LGM, 26.5-20 ka) (Clark et al., 2009). The MAT of MIS 2, including LGM, was relatively stable. The temperature increased again from late MIS 2 towards MIS 1. Compared to the east side of the Korean Peninsula, the west side tends to be relatively warmer. The range of MAT at the sites over our study period is 2-10℃. Gihwari and Gigok sites have the lowest (3℃) and Jeungsan-ri site has the highest MAT (8℃). The mean MAT for all sites fluctuated within 4℃ between 50 ka and 10 ka. The first appearance of stemmed points occurred in the middle of the decreasing MAT trend in MIS 3, at 40-35 ka. Figure 7D shows a negative relationship between temperature and elevation. For example, Gihwari Cave site, with one of the lowest MAT distributions, is located in the highest elevation around 300m above sea level. Jeungsanri is located at a lower elevation and has one of the highest MAT distributions.

@@

Figure \@ref(fig:spd-model) shows the summed probability distribution (SPD) of the radiocarbon ages from the sites in our sample overlaid with distributions (indicated by the grey envelopes) generated from four models of demographic change for the period 50 ka to 10 ka (exponential, uniform, linear, and logistic). The fit of models to our observed SPD was computed using 1000 simulations, and the resulting global p-values indicate significant deviations in the observed SPD from the modeled distributions. Overall, the linear model is the best fit with the observed SPD, having the lowest AIC value of the three models where we could compute this. We can say that the population did not grow or decline strongly logistically, exponentially or linearly during the Late Pleistocene, but instead followed a noisy distribution.  The most striking results here are the SPD showing a steady growth rising to a peak at about 35 ka, indicated by the significant deviations from all four null models (shaded in red). The SPD remains high with minor deviations until declining to a lower level at 32 ka. After 32 ka the SPD remains generally low, with frequent temporary fluctuations.

```{r spd-model, fig.cap=paste0("Summed probability distribution of ", nrow(dates_clean), " radiocarbon dates. The black solid line represents actual radiocarbon ages. The grey shaded region shows the Monte Carlo envelope that encompasses the 95% confidence interval for the null models. The red and blue vertical bands highlight the portions of the SPD where positive and negative deviations are detected. We applied logistic, exponential and linear to AIC along with the logistical growth model (fit) to evaluate each model [@riris2019widespread]. We excluded the uniform model because it has no fitted curve or surface to evaluate. The best fit model has the lowest AIC score. In this case, the linear model is the best fit for the observed data"), cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/009-summed-probability-distributions.R"))}
quietly(source(here::here("analysis/paper/009-summed-probability-distributions.R")))
knitr::include_graphics(here::here("analysis/figures/008-summed-probability-distribution-models.png"))
```
